A Toolbox for Manipulating and Assessing Color Palettes for Statistical Graphics

### 2019-02-24 pkg/colorspace/paper/Toolbox/article.Rmd
removed some details and redundancies
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@454 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

```{r preliminaries, echo=FALSE, message=FALSE}
library("colorspace")
library("ggplot2")
theme_set(theme_minimal())
```

@@

```{r hcl-properties, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 4, fig.height = 2.2, out.width = "0.5\\textwidth", fig.cap = "Axes of the HCL color space: Hue (= type of color, dominant wavelength), chroma (= colorfulness), luminance (= brightness). In each line the color coordinates are varied along the indicated axis, keeping the other two coordinates fixed."}
swatchplot(
  "Hue"       = sequential_hcl(5, h = c(0, 300), c = c(60, 60), l = 65),
  "Chroma"    = sequential_hcl(5, h = 0, c = c(100, 0), l = 65, rev = TRUE, power = 1),
  "Luminance" = sequential_hcl(5, h = 260, c = c(25, 25), l = c(25, 90), rev = TRUE, power = 1),
  off = 0
)
```

@@

## Installation

@@

```{r hcl-palettes, message = FALSE, warning = FALSE, fig.height = 6.75, fig.width = 12, out.width = "\\textwidth", fig.cap = "Overview of available predefined palettes in \\pkg{colorspace}."}
library("colorspace")
hcl_palettes(plot = TRUE)
```

@@

```{r eustockmarkets-plot, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Time series plot of \\code{EuStockMarkets} data with \\code{qualitative\\_hcl(4, \"Dark 3\")} palette."}
q4 <- qualitative_hcl(4)
par(mar = c(5, 4, 1, 1))
plot(log(EuStockMarkets), plot.type = "single", col = q4, lwd = 2)
legend("topleft", colnames(EuStockMarkets), col = q4, lwd = 3, bty = "n")
```

@@

```{r titanic-plot, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
ttnc <- margin.table(Titanic, c(1, 4))[, 2:1]
par(mar = c(5, 4, 1, 1))
spineplot(ttnc, col = sequential_hcl(2, "Purples 3"))
```

@@

To plug the HCL color palettes into \pkg{ggplot2} graphics suitable discrete and/or
continuous \pkg{gglot2} color scales are provided. The scales are called via the scheme
`scale_<aesthetic>_<datatype>_<colorscale>()`, where `<aesthetic>` is the name
of the aesthetic (`fill`, `color`, `colour`), `<datatype>` is the type of the
variable plotted (`discrete` or `continuous`) and `<colorscale>` sets the type
of the color scale used (`qualitative`, `sequential`, `diverging`,
`divergingx`).

@@

```{r iris-ggplot, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
library("ggplot2")
ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) +
  scale_fill_discrete_qualitative(palette = "Dark 3")
```

@@

```{r diamonds-ggplot, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
dsamp <- diamonds[1 + 1:1000 * 50, ]
ggplot(dsamp, aes(carat, price, color = cut)) + geom_point() +
  scale_color_discrete_sequential(palette = "Purples 3", nmax = 6, order = 2:6)
```

@@

As motivated in the previous article ([Color Spaces: S4 Classes and Utilities](color_spaces.html)),
the HCL space is particularly useful for specifying individual colors and color palettes as its three axes match those
of the human visual system very well. Therefore, the \pkg{colorspace} package provides
three types of palettes based on the HCL model:

* _Qualitative:_ Designed for coding categorical information, i.e.,
  where no particular ordering of categories is available and every color
  should receive the same perceptual weight.
* _Sequential:_ Designed for coding ordered/numeric information, i.e.,
  going from high to low (or vice versa).
* _Diverging:_ Designed for coding ordered/numeric information around a central
  neutral value, i.e., where colors diverge from neutral to two extremes.

The corresponding functions are `qualitative_hcl()`, `sequential_hcl()`,
and `diverging_hcl()`. Their construction principles are exemplified in the
following color swatches and explained in more detail below. The desaturated
palettes bring out clearly that luminance differences (= light-dark contrasts),
or the lack thereof, are crucial for coding the different kinds of palettes.

@@

@@

# HCL-based color scales for ggplot2 {#sec:ggplot2_color_scales}

All [HCL-based color palettes](hcl_palettes.html) in the \pkg{colorspace} package [@color:colorspace]
are also provided as discrete and/or continuous color scales for the use with the _ggplot2_ package
[@color:ggplot2; @color:ggplot2pkg].

The scales are called via the scheme

A few examples of these scales are illustrated in the following sections.

## Using the scales in ggplot2

A discrete qualitative scale applied to a fill aesthetic corresponds to the function `scale_fill_discrete_qualitative()`:

```{r density-qualitative, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) +
  scale_fill_discrete_qualitative()
```

Similarly, a color aesthetic for a discrete qualitative scale corresponds to the function `scale_color_discrete_qualitative()`:

```{r point-qualitative, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + geom_point() +
  scale_color_discrete_qualitative(palette = "Set 2")
```

A continuous sequential scale applied to a color aesthetic corresponds to the function `scale_color_continuous_sequential()`:

```{r jitter-sequential, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
ggplot(iris, aes(x = Species, y = Sepal.Width, color = Sepal.Length)) + geom_jitter(width = 0.2) +
  scale_color_continuous_sequential(palette = "Heat")
```

A continuous sequential scale applied to a fill aesthetic corresponds to the function `scale_fill_continuous_sequential()`:

```{r raster-sequential, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
df <- data.frame(height = c(volcano), x = c(row(volcano)), y = c(col(volcano)))
ggplot(df, aes(x, y, fill = height)) + 
  geom_raster() + 
  coord_fixed(expand = FALSE) +
  scale_fill_continuous_sequential(palette = "Blues")
```

A continuous diverging scale applied to a fill aesthetic corresponds to the function `scale_fill_continuous_diverging()`:

```{r tile-diverging, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
cm <- cor(mtcars)
df <- data.frame(cor = c(cm), var1 = factor(col(cm)), var2 = factor(row(cm)))
levels(df$var1) <- levels(df$var2) <- names(mtcars)
ggplot(df, aes(var1, var2, fill = cor)) + 
  geom_tile() + 
  coord_fixed() +
  ylab("variable") +
  scale_x_discrete(position = "top", name = "variable") +
  scale_fill_continuous_diverging("Blue-Red 3")
```

## Customizing the scales

All scale functions accept a `palette` argument which allows you to pick a specific color palette out of a selection of different options. All available palettes are listed at the end of this document. For example, we could use the "Harmonic" palette when we need a qualitative color scale:

```{r density-qualitative2, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) +
  scale_fill_discrete_qualitative(palette = "Harmonic")
```

The color palettes are calculated on the fly depending on the number of different colors needed. But sometimes, it may be desireable to pick specific colors out of a larger set, e.g., when we are making two plots where one contains a subset of the data of the other, or when the default order of colors is not ideal. Therefore, all discrete scales provide parameters `nmax` to set the total number of colors requested and `order` (a vector of integers) to reorder the color palette.

Applied to the previous plot, we could for example do the following:

```{r density-qualitative3, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) +
  scale_fill_discrete_qualitative(palette = "Harmonic", nmax = 5, order = c(5, 1, 2))
```

The `nmax` option is also convenient to remove some colors from a scale that may not be appropriate for the plot. For example, the `scale_color_brewer()` scale that comes with _ggplot2_ tends to produce points that are too light:

```{r point-brewer, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
dsamp <- diamonds[1 + 1:1000 * 50, ]
gg <- ggplot(dsamp, aes(carat, price, color = cut)) + geom_point()
gg + scale_color_brewer(palette = "Blues")
```

Similar problems can arise with the HCL palettes, but there we have the option of creating additional colors that we then do not use:

```{r point-sequential, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
gg + scale_color_discrete_sequential(palette = "Blues", nmax = 6, order = 2:6)
```

(We use `order = 2:6` to pick the five darkest colors and omit the lightest color.)

All [HCL-based color palettes](hcl_palettes.html) are defined via sets of hue (H), chroma (C), and luminance (L) values. For example, the qualitative scales vary hue from a start value to an end value while keeping chroma and luminance fixed. Similarly, single-hue sequential scales vary chroma and luminance while keeping the hue fixed. We can override these settings by specifying the corresponding H, C, or L values in addition to the palette name. As an example, consider the following plot:

```{r jitter-sequential2, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
ggplot(iris, aes(x = Species, y = Sepal.Width, color = Sepal.Length)) + geom_jitter(width = 0.2) +
  scale_color_continuous_sequential(palette = "Terrain")
```

Now assume we generally like the color scale but find it a bit too pink at the end. We can fix this issue by specifying an alternative final hue value, e.g., `h2 = 60` corresponding to yellow:

```{r jitter-sequential3, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
ggplot(iris, aes(x = Species, y = Sepal.Width, color = Sepal.Length)) + geom_jitter(width = 0.2) +
  scale_color_continuous_sequential(palette = "Terrain", h2 = 60)
```

The next example uses a diverging scale. First consider the plot with the unmodified "Blue-Yellow 2" palette:

```{r tile-diverging2, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
cm <- cor(mtcars)
df <- data.frame(cor = c(cm), var1 = factor(col(cm)), var2 = factor(row(cm)))
levels(df$var1) <- levels(df$var2) <- names(mtcars)
gg <- ggplot(df, aes(var1, var2, fill = cor)) + 
  geom_tile() + 
  coord_fixed() +
  ylab("variable") +
  scale_x_discrete(position = "top", name = "variable")

gg + scale_fill_continuous_diverging(palette = "Blue-Yellow 2")
```

And now the same plot with some palette customizations: The ordering is reversed so that blue is used for
positive correlations and yellow for negative ones. Moreover, the power parameter `p2` for the luminance
is increased so that only correlations close to an absolute value of 1 have dark colors while intermediate
correlations have relatively light colors.

```{r tile-diverging3, fig.width = 6.5, fig.asp = .7, fig.cap = "Foo."}
gg + scale_fill_continuous_diverging(palette = "Blue-Yellow 2", rev = TRUE, p2 = 2)
```

See the reference manual for the exact set of customization parameters that are available for each scale.

The continuous scales also provide the option to limit the scale range to which data are mapped, via the parameters `begin` and `end`. As an example, assume we are using the approximation of the viridis scale provided by `scale_color_continuous_sequential()`:

```{r jitter-sequential4, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
ggplot(iris, aes(x = Species, y = Sepal.Width, color = Sepal.Length)) + geom_jitter(width = 0.2) +
  scale_color_continuous_sequential(palette = "Viridis")
```

If we want to remove some of the darkest blues and some of the brightest yellows from this scale, we can write:

```{r jitter-sequential5, fig.width = 6.5, fig.asp = 0.7, fig.cap = "Foo."}
ggplot(iris, aes(x = Species, y = Sepal.Width, color = Sepal.Length)) + geom_jitter(width = 0.2) +
  scale_color_continuous_sequential(palette = "Viridis", begin = 0.15, end = 0.9)
```

## Available palettes

In the following, we are visualizing all scales currently available via
pre-defined names. These visualizations are generated by the function
`hcl_palettes()` with option `plot = TRUE`.

The discrete qualitative scales are all called via `scale_*_discrete_qualitative(palette = "name")`,
where `name` is the name of the palette, e.g., `palette = "Pastel 1"`. There are no continuous qualitative
scales.

The discrete sequential scales are all called via `scale_*_discrete_sequential(palette = "name")`,
where `name` is the name of the palette, e.g., `palette = "Grays"`. Continuous approximations to the discrete
scales exist and can be called via `scale_*_continuous_sequential(palette = "name")`

The discrete diverging scales are all called via `scale_*_discrete_diverging(palette = "name")`,
where `name` is the name of the palette, e.g., `palette = "Blue-Red"`. Continuous approximations to the discrete
scales exist and can be called via `scale_*_continuous_diverging(palette = "name")`

@@

As a first simple example, the properties of the three HCL axes are brought out by varying
only one coordinate while keeping the other two fixed:

```{r swatch-hcl, fig.width = 4, fig.height = 2.2, fig.cap = "Foo."}
swatchplot(
  "Hue"       = sequential_hcl(5, h = c(0, 300), c = c(60, 60), l = 65),
  "Chroma"    = sequential_hcl(5, h = 0, c = c(100, 0), l = 65, rev = TRUE, power = 1),
  "Luminance" = sequential_hcl(5, h = 260, c = c(25, 25), l = c(25, 90), rev = TRUE, power = 1),
  off = 0
)
```

This shows the following:

- Hue: Only the hue (= type of color) changes from H = 0 (red) via 60 (yellow), etc. to 300 (purple)
  while chroma and luminance are fixed to moderate values of C = 60 and L = 65, respectively.
- Chroma: Only the chroma (= colorfulness) changes from C = 0 (gray) to 100 (colorful)
  while hue and luminance are fixed to H = 0 (red) and L = 65, respectively.
- Luminance: Only the luminance (= brightness) changes from L = 90 (light) to 25 (dark)
  while hue and chroma are fixed to H = 260 (blue) and C = 25 (low, close to gray), respectively.

@@

```{r spec-qualitative, fig.width = 5, fig.height = 5, fig.cap = "Foo."}
specplot(qualitative_hcl(100, "Set 2"))
```

@@

```{r spec-sequential, fig.width = 5, fig.height = 5, fig.cap = "Foo."}
specplot(sequential_hcl(100, "Blues 2"))
```

@@

```{r spec-diverging, fig.width = 5, fig.height = 5, fig.cap = "Foo."}
specplot(diverging_hcl(100, "Blue-Red"))
```

@@

```{r spec-rainbow, fig.width = 5, fig.height = 7.5, fig.cap = "Foo."}
specplot(rainbow(100), rgb = TRUE)
```

@@

This GUI can either be started with `hclwizard()` (or equivalently `hcl_wizard()`) which
by default starts the _shiny_ version:

![](../../vignettes/screenshots/choose_palette_1.png)

The _tcltk_ version is started by default with `choose_palette()`:

![](../../vignettes/screenshots/choose_palette_3.png)

However, all defaults can be modified by setting  `gui = "tcltk"` or `gui = "shiny"`.

The GUIs interface the palette functions `qualitative_hcl()` for qualitative palettes,
`sequential_hcl()` for sequential palettes with single or
multiple hues, and `diverging_hcl()` for diverging palettes (composed
from two single-hue sequential palettes). See the discussion of
[HCL-based color palettes](hcl_palettes.html) for more details.

Optionally, the active palette can be illustrated by using a
[range of examples](palette_visualization.html#demonstration-of-statistical-graphics)
such as a map, heatmap, scatter plot, perspective 3D surface etc.
To demonstrate different types of deficiencies, the active palette may be
desaturated (emulating printing on a grayscale printer) and collapsed to
emulate different types of color-blindness (without red-green or green-blue
contrasts) using the `simulate_cvd()` functions. To facilitate generation of
palettes for black/dark backgrounds, a "dark mode" of the GUIs is also available:

![](../../vignettes/screenshots/choose_palette_2.png)

![](../../vignettes/screenshots/choose_color.png)

![](../../vignettes/screenshots/cvd_emulator_1.png){width=49%}
![](../../vignettes/screenshots/cvd_emulator_2.png){width=49%}

It allows to upload a raster image in
JPG or PNG format which is then check for various kinds of color vision deficiencies
at the selected severity. By default the severity is set to 100% and all supported
kinds of color vision deficiency are checked for, i.e.,

* Monochromatic (desaturated grayscale).
* Deuteranope vision (green deficient).
* Protanope vision (red deficient).
* Tritanope vision (blue deficient).

@@

```{r volcano, echo = FALSE, fig.height = 3, fig.width = 5.5, dev = "png", fig.cap = "Foo.", dpi = 100}
z <- 2 * volcano
x <- 10 * (1:nrow(z))
y <- 10 * (1:ncol(z))
par(mar = rep(0, 4))
persp(x, y, z, theta = 10, phi = 30, col = "lightgray", scale = FALSE,
  ltheta = 100, shade = 0.5, border = NA, box = FALSE)
```

@@

```{r heatmap-sequential, echo = FALSE, fig.height = 15, fig.width = 9, out.width = "\\textwidth", fig.cap = "Foo."}
par(mfrow = c(5, 2), mar = rep(0.5, 4), oma = c(1, 7.9, 3, 1), las = 1)
rain <- rainbow(11, end = 2/3)
blyl <- sequential_hcl(11, "Blue-Yellow")
demoplot(rain, "heatmap")
mtext("Original", side = 2, line = 1)
mtext("rainbow(11, end = 2/3)", line = 1)
demoplot(blyl, "heatmap")
mtext("sequential_hcl(11, \"Blue-Yellow\")", line = 1)
demoplot(desaturate(rain), "heatmap")
mtext("Desaturated", side = 2, line = 1)
demoplot(desaturate(blyl), "heatmap")
demoplot(deutan(rain), "heatmap")
mtext("Deuteranope", side = 2, line = 1)
demoplot(deutan(blyl), "heatmap")
demoplot(protan(rain), "heatmap")
mtext("Protanope", side = 2, line = 1)
demoplot(protan(blyl), "heatmap")
demoplot(tritan(rain), "heatmap")
mtext("Tritanope", side = 2, line = 1)
demoplot(tritan(blyl), "heatmap")
```

@@

## Illustration: Map with diverging palette

As another example for the poor performance of the RGB rainbow we employ a
shaded map. This is available as `demoplot(..., "map")` and is based on county
polygons for Alabama, Georgia, and South Carolina along with an artifical variable
used for coloring.

Often the red-yellow-green RGB spectrum is used for a diverging palette with 
yellow as the neutral value. In base R this can be generated by
`rainbow(11, end = 1/3)`. However, this palette has again a number of weaknesses,
especially that the green-yellow part of the palette almost collapses to the same
color when desaturated or when color blindness is emulated.

To illustrate that much more balanced palettes for the same purpose are available
the `Geyser` palette [mimicked from @color:CARTO] is adopted:
`divergingx_hcl(11, "Geyser", rev = TRUE)`. This also goes from red via yellow to
a green/blue and hence needs the `divergingx_hcl()` function (as opposed to
`diverging_hcl()` which always uses a neutral gray).

```{r map-diverging, echo = FALSE, fig.height = 13, fig.width = 9, out.width = "\\textwidth", fig.cap = "Foo."}
par(mfrow = c(5, 2), mar = rep(0.5, 4), oma = c(1, 7.9, 3, 1), las = 1)
rain <- rainbow(11, end = 1/3)
gysr <- divergingx_hcl(11, "Geyser", rev = TRUE)
demoplot(rain, "map")
mtext("Original", side = 2, line = 1)
mtext("rainbow(11, end = 1/3)", line = 1)
demoplot(gysr, "map")
mtext("divergingx_hcl(11, \"Geyser\", rev = TRUE)", line = 1)
demoplot(desaturate(rain), "map")
mtext("Desaturated", side = 2, line = 1)
demoplot(desaturate(gysr), "map")
demoplot(deutan(rain), "map")
mtext("Deuteranope", side = 2, line = 1)
demoplot(deutan(gysr), "map")
demoplot(protan(rain), "map")
mtext("Protanope", side = 2, line = 1)
demoplot(protan(gysr), "map")
demoplot(tritan(rain), "map")
mtext("Tritanope", side = 2, line = 1)
demoplot(tritan(gysr), "map")
```

While many versions of the RGB rainbow displays are hard to read because
they do not bring out any differences in the green-yellow arm of the palette,
the HCL-based palette works reasonably well in all settings. Only the grayscale
version cannot bring out the different arms of the palette. However, at least
both directions of deviation are visible even if they cannot be distinguished.
This is preferable to the RGB rainbow which hides all differences in the
green-yellow arm of the palette. (However, if grayscale printing is desired
a sequential rather than a diverging palette is probably necessary.)

@@

```{r desaturate-wheel, fig.height = 6, fig.width = 6, fig.cap = "Foo."}
wheel <- function(col, radius = 1, ...)
  pie(rep(1, length(col)), col = col, radius = radius, ...) 
par(mar = rep(0.5, 4), mfrow = c(2, 2))
wheel(rainbow(8))
wheel(desaturate(rainbow(8)))
wheel(rainbow_hcl(8))
wheel(desaturate(rainbow_hcl(8)))
```

@@

```{r lighten-darken, fig.height = 3.5, fig.width = 4.5, fig.cap = "Foo."}
oi <- c("#61A9D9", "#ADD668", "#E6D152", "#CE6BAF", "#797CBA")
swatchplot(
  "-40%" = lighten(oi, 0.4),
  "-20%" = lighten(oi, 0.2),
  "  0%" = oi,
  " 20%" =  darken(oi, 0.2),
  " 40%" =  darken(oi, 0.4),
  off = c(0, 0)
)
```

@@

In the plots below more combinations are visualized: In the left panel for maximum
chroma across hues given luminance and in the right panel with increasing luminance
given hue.

```{r max-chroma-plot, echo = FALSE, fig.height = 4.5, fig.width = 10, out.width = "\\textwidth", fig.cap = "Foo."}
H <- seq(0, 360, by = 10)
L <- seq(30, 90, by = 20)
C <- outer(H, L, max_chroma)

par(mfrow = c(1, 2), xaxs = "i", yaxs = "i", mar = c(5, 4, 1, 5), las = 1)
plot(0, 0, type = "n", xlim = c(0, 360), ylim = c(0, 165),
  xlab = "Hue (H)", ylab = "Maximum chroma (C)")
axis(4, at = C[37, ], labels = paste("L =", L))
for(i in seq_along(L)) {
  lines(H, C[, i])
  points(H, C[, i], col = hex(polarLUV(L[i], C[, i], H)), pch = 19, cex = 1.5, xpd = TRUE)
}

L <- seq(0, 100, by = 5)
H <- c(0, 60, 120, 250, 330)
C <- outer(H, L, max_chroma)

par(xaxs = "i", yaxs = "i", mar = c(5, 4, 1, 1), las = 1)
plot(0, 0, type = "n", xlim = c(0, 175), ylim = c(0, 100),
  xlab = "Maximum chroma (C)", ylab = "Luminance (L)")
for(i in rev(seq_along(H))) {
  lines(C[i, ], L)
  points(C[i, ], L, col = hex(polarLUV(L, C[i, ], H[i])), pch = 19, cex = 1.5, xpd = TRUE)
  j <- which.max(C[i,])
  text(C[i, j], L[j], paste("H =", H[i]), pos = 4, col = hex(polarLUV(L[j], C[i, j], H[i])))
}
points(0, 100, cex = 1.6, xpd = TRUE)
```

@@

```{r mixcolor, fig.height = 1.5, fig.width = 5, fig.cap = "Foo."}
R <- RGB(1, 0, 0)
G <- RGB(0, 1, 0)
Y <- mixcolor(0.5, R, G)
Y
par(mar = rep(0, 4))
swatchplot(c(hex(R), hex(Y), hex(G)))
```

# Approximating palettes from other packages {#sec:approximations}

## Approximations of ColorBrewer.org palettes

```{r brewer, results = "hide", message = FALSE, warning = FALSE, fig.width = 6, fig.height = 6, fig.cap = "Foo."}
specplot(
  rev(RColorBrewer::brewer.pal(9, "YlGnBu")),
  sequential_hcl(9, "YlGnBu"),
  main = "ColorBrewer.org: YlGnBu"
)
```

```{r viridis, results = "hide", message = FALSE, warning = FALSE, fig.width = 6, fig.height = 6, fig.cap = "Foo."}
specplot(
  viridis::viridis(9),
  sequential_hcl(9, "Viridis"),
  main = "Viridis"
)
```

>>>>

```{r preliminaries, echo = FALSE, message = FALSE}
library("colorspace")
library("ggplot2")
theme_set(theme_minimal())
knitr::opts_chunk$set(message = FALSE, warning = FALSE,
  dev = "pdf", dpi = 150,
  fig.height = 6, fig.width = 6, fig.pos = "t!",
  fig.align = "center", fig.show = "hold", fig.path = "Figures/")
```

@@

```{r hcl-properties, echo = FALSE, fig.width = 4, fig.height = 2.2, out.width = "0.5\\textwidth", fig.cap = "Axes of the HCL color space. Top: Hue $H$ changes from 0 (red) via 60 (yellow), etc.\\ to 300 (purple) with fixed $C = 60$ and $L = 65$. Center: Chroma $C$ changes from 0 (gray) to 100 (colorful) with fixed $H = 0$ (red) and $L = 65$. Bottom: Luminance $L$ changes from 90 (light) to 25 (dark) with fixed $H = 260$ (blue) and $C = 25$ (low, close to gray)."}
swatchplot(
  "Hue\n(Type of color)"       = sequential_hcl(5, h = c(0, 300), c = c(60, 60), l = 65),
  "Chroma\n(Colorfulness)"    = sequential_hcl(5, h = 0, c = c(100, 0), l = 65, rev = TRUE, power = 1),
  "Luminance\n(Brightness)" = sequential_hcl(5, h = 260, c = c(25, 25), l = c(25, 90), rev = TRUE, power = 1),
  nrow = 3, line = 5.2, off = 0
)
```

@@

@@

```{r hcl-palettes, fig.height = 6.75, fig.width = 12, out.width = "\\textwidth", fig.cap = "Overview of available predefined palettes in \\pkg{colorspace}."}
library("colorspace")
hcl_palettes(plot = TRUE)
```

@@

```{r eustockmarkets-plot, echo = FALSE, warning = FALSE, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Time series plot of \\code{EuStockMarkets} data with \\code{qualitative\\_hcl(4, \"Dark 3\")} palette."}
q4 <- qualitative_hcl(4)
par(mar = c(5, 4, 1, 1))
plot(log(EuStockMarkets), plot.type = "single", col = q4, lwd = 2)
legend("topleft", colnames(EuStockMarkets), col = q4, lwd = 3, bty = "n")
```

@@

```{r titanic-plot, echo = FALSE, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
ttnc <- margin.table(Titanic, c(1, 4))[, 2:1]
par(mar = c(5, 4, 1, 1))
spineplot(ttnc, col = sequential_hcl(2, "Purples 3"))
```

@@

To plug the HCL color palettes into \pkg{ggplot2} graphics [@color:ggplot2; @color:ggplot2pkg]
suitable discrete and/or continuous \pkg{gglot2} color scales are provided.
The scales are called via the scheme

@@

```{r iris-ggplot, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
library("ggplot2")
ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) +
  scale_fill_discrete_qualitative(palette = "Dark 3")
```

@@

```{r diamonds-ggplot, fig.height = 4, fig.width = 6, out.width = "0.75\\textwidth", fig.cap = "Foo."}
dsamp <- diamonds[1 + 1:1000 * 50, ]
ggplot(dsamp, aes(carat, price, color = cut)) + geom_point() +
  scale_color_discrete_sequential(palette = "Purples 3", nmax = 6, order = 2:6)
```

@@

As motivated in the previous section,
the HCL space is particularly useful for specifying individual colors and color palettes as its three axes match those
of the human visual system very well. Therefore, the \pkg{colorspace} package provides
three palette functions based on the HCL model: `qualitative_hcl()`, `sequential_hcl()`,
and `diverging_hcl()`. Their construction principles are exemplified in
Figure\ \ref{fig:hcl-palettes-principles} and explained in more detail below. The desaturated
palettes bring out clearly that luminance differences (= light-dark contrasts),
or the lack thereof, are crucial for coding the different kinds of palettes.

@@

## Approximating palettes from other packages

```{r brewer-carto-viridis, results = "hide", echo = FALSE, fig.width = 6, fig.height = 6, out.width = "0.49\\textwidth", fig.cap = "Foo."}
specplot(
  rev(RColorBrewer::brewer.pal(7, "YlGnBu")),
  sequential_hcl(7, "YlGnBu"),
  main = "ColorBrewer.org: YlGnBu"
)
specplot(
  viridis::viridis(7),
  sequential_hcl(7, "Viridis"),
  main = "viridis: Viridis"
)
specplot(
  rcartocolor::carto_pal(7, "ag_Sunset"),
  sequential_hcl(7, "ag_Sunset"),
  main = "CARTO: ag_Sunset"
)
specplot(
  viridis::plasma(7),
  sequential_hcl(7, "Plasma"),
  main = "viridis: Plasma"
)
```

@@

@@

As a first simple example:
`swatchplot("Label" = sequential_hcl(5))`.

@@

```{r spec-qualitative, fig.width = 5, fig.height = 5, out.width = "0.5\\textwidth", fig.cap = "Foo."}
specplot(qualitative_hcl(100, "Set 2"))
```

@@

```{r spec-sequential, fig.width = 5, fig.height = 5, out.width = "0.5\\textwidth", fig.cap = "Foo."}
specplot(sequential_hcl(100, "Blues 2"))
```

@@

```{r spec-diverging, fig.width = 5, fig.height = 5, out.width = "0.5\\textwidth", fig.cap = "Foo."}
specplot(diverging_hcl(100, "Blue-Red"))
```

@@

```{r spec-rainbow, fig.width = 5, fig.height = 7.5, out.width = "0.5\\textwidth", fig.cap = "Foo."}
specplot(rainbow(100), rgb = TRUE)
```

@@

@@

```{r volcano, echo = FALSE, fig.height = 3, fig.width = 5.5, dev = "png", fig.cap = "Foo.", dpi = 100, out.width = "0.5\\textwidth"}
z <- 2 * volcano
x <- 10 * (1:nrow(z))
y <- 10 * (1:ncol(z))
par(mar = rep(0, 4))
persp(x, y, z, theta = 10, phi = 30, col = "lightgray", scale = FALSE,
  ltheta = 100, shade = 0.5, border = NA, box = FALSE)
```

@@

```{r heatmap-sequential, echo = FALSE, fig.height = 15, fig.width = 9, out.width = "0.85\\textwidth", fig.pos = "p!", fig.cap = "Foo."}
par(mfrow = c(5, 2), mar = rep(0.5, 4), oma = c(1, 7.9, 3, 1), las = 1)
rain <- rainbow(11, end = 2/3)
blyl <- sequential_hcl(11, "Blue-Yellow")
demoplot(rain, "heatmap")
mtext("Original", side = 2, line = 1)
mtext("rainbow(11, end = 2/3)", line = 1)
demoplot(blyl, "heatmap")
mtext("sequential_hcl(11, \"Blue-Yellow\")", line = 1)
demoplot(desaturate(rain), "heatmap")
mtext("Desaturated", side = 2, line = 1)
demoplot(desaturate(blyl), "heatmap")
demoplot(deutan(rain), "heatmap")
mtext("Deuteranope", side = 2, line = 1)
demoplot(deutan(blyl), "heatmap")
demoplot(protan(rain), "heatmap")
mtext("Protanope", side = 2, line = 1)
demoplot(protan(blyl), "heatmap")
demoplot(tritan(rain), "heatmap")
mtext("Tritanope", side = 2, line = 1)
demoplot(tritan(blyl), "heatmap")
```

@@

This GUI can either be started with `hclwizard()` (or equivalently `hcl_wizard()`) which
by default starts the _shiny_ version.
The _tcltk_ version is started by default with `choose_palette()`.
However, all defaults can be modified by setting  `gui = "tcltk"` or `gui = "shiny"`.

```{r choose_palette, echo = FALSE, out.width = "\\textwidth", fig.cap="Foo."}
knitr::include_graphics("../../vignettes/screenshots/choose_palette_1.png")
```

The GUIs interface the palette functions `qualitative_hcl()` for qualitative palettes,
`sequential_hcl()` for sequential palettes with single or
multiple hues, and `diverging_hcl()` for diverging palettes (composed
from two single-hue sequential palettes).

Optionally, the active palette can be illustrated by using a
[range of examples](palette_visualization.html#demonstration-of-statistical-graphics)
such as a map, heatmap, scatter plot, perspective 3D surface etc.
To demonstrate different types of deficiencies, the active palette may be
desaturated (emulating printing on a grayscale printer) and collapsed to
emulate different types of color-blindness (without red-green or green-blue
contrasts) using the `simulate_cvd()` functions. To facilitate generation of
palettes for black/dark backgrounds, a "dark mode" of the GUIs is also available.

```{r choose_color, echo = FALSE, out.width = "\\textwidth", fig.cap="Foo."}
knitr::include_graphics("../../vignettes/screenshots/choose_color.png")
```

\begin{figure}[t!]
\centering 
\includegraphics[width=0.49\textwidth]{../../vignettes/screenshots/cvd_emulator_1} 
\includegraphics[width=0.49\textwidth]{../../vignettes/screenshots/cvd_emulator_2}
\caption[Foo]{Foo.}\label{fig:cvd_emulator}
\end{figure}

It allows to upload a raster image in
JPG or PNG format which is then checked for various kinds of color vision deficiencies
at the selected severity. By default the severity is set to 100% and all supported
kinds of color vision deficiency are checked for.

@@

```{r desaturate-wheel, out.width = "0.5\\textwidth", fig.cap = "Foo."}
wheel <- function(col, radius = 1, ...)
  pie(rep(1, length(col)), col = col, radius = radius, ...) 
par(mar = rep(0.5, 4), mfrow = c(2, 2))
wheel(rainbow(8))
wheel(desaturate(rainbow(8)))
wheel(rainbow_hcl(8))
wheel(desaturate(rainbow_hcl(8)))
```

@@

```{r lighten-darken, fig.height = 3.5, fig.width = 4.5, out.width = "0.5\\textwidth", fig.cap = "Foo."}
oi <- c("#61A9D9", "#ADD668", "#E6D152", "#CE6BAF", "#797CBA")
swatchplot(
  "-40%" = lighten(oi, 0.4),
  "-20%" = lighten(oi, 0.2),
  "  0%" = oi,
  " 20%" =  darken(oi, 0.2),
  " 40%" =  darken(oi, 0.4),
  off = c(0, 0)
)
```

@@

@@

```{r mixcolor}
R <- RGB(1, 0, 0)
G <- RGB(0, 1, 0)
Y <- mixcolor(0.5, R, G)
Y
```
